Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-29 doc/report/report.tex
report

\subsection{Swedish}
Swedish\cite{SAG:intro} is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite{H&H:1027}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=60mm]{gfTree.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

unless their in passive form
\enumsentence{
Det äts en fågel på taket \\
A bird is being eaten on the roof}
A very common example of this is sentences with the verbs \emph{finnas},
\emph{saknas} and \emph{fattas}.
\enumsentence{\shortex{3}
{Det & finns & kaffe}
{It & exist & coffee}
{''There is coffee"}}
\enumsentence{\shortex{3}
{Det & saknas&  kaffe}
{It & misses & coffee}
{''There is no coffee"}}
\enumsentence{\shortex{3}
{Det & fattas & kaffe}
{It & misses & coffee}
{''There is no coffee"}}
A special GF category was needed for this, the \verb|SimpleVP|.

There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågel på taket. & **Det sitter fåglarna på taket.\\
There are that bird sitting on the roof & There are the birds sitting on the roof \\
\end{tabular}}
The distinction lays in the definitness of the noun: 
it must be indefinite, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determinened form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definitness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers')

finnas: här får man ha definit: det finns de som inte vill. Samma RelCl som
jag just la till, de äppplen. (de + relsats)

'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'. Kan säga 'en sån katt som sover, med 'sådan' + Rel
%har lagt till Det äpple + Rel (RelCNNP). *Det äpple är rött. men Det äpple jag äter är rött.
%

\subsubsection{Quantifiers}
pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
has to decide, one category for each?
Genitive -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form
           fixing this include restructuring all
Det-utr. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

 'de flesta de gula hästarna'

-- or is this really just AdVs? 
to change position of adv:
normal: after verb 'han äpplet redan'
now want: 'äter redan äpplet'
          'har redan ätit äpplet' 
          'har ätit redan äpplet' :( which doesn't work either :)

but have other words like 'bara' which can be put in other positions.
  but needs to restruct whole of VP, since no field is befor fin.
  Can also act before the noun or as AdV. Examples.
\verb|s SPres Simul Neg Main : han bara sover inte alltid|
  have added a field in VP, a0, show table.
  makes some weird constructions when using modal verbs, 
  \verb|s SPres Anter Pos Main : han bara har alltid sovit|
  but rather than changing to 'han har bara sovit', we keep it. Is a possible
  tolkning and the other case is covered by using it as an AdV, we don't want
  ambiguities.

Also needed categories for 'pronouns' wich can act in different ways.
and 'antal','sorter' - need indetermined, antal plural, sorter either.
Should be able to modify by a godtyckligt antal adjective, and they must be determined.
'(ett stort antal) människor' and not 'ett stort (antal människor)'
Don't want them as a NPs or CNs since we don't want other Ns to be used this way
'en flicka kaffe', this is then Appos? (Or do we?). Should also not be utsatta for what other
NPs or CN can be. Dessutom, we want to make use of the comp-fiel. 'mamma till', but is this really
the same??

The VP require many changes, to add passive (no), PPart, bara.
About PPart, but do we need it with new Saldo? 'de är stuckna' , 'den nyligen funna'
for the last one, we need VPSlash to allow adverb.

%% Refl 
\subsubsection*{Reflexive pronouns}
Could varandra be thought of as reflexive for 'oss'?
varandra and sin/sitt/sig. Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.

@@

Kanske mindre om detta, skriv om vad jag nu la till.
om 'de äpplen'
Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\

>>>>

\subsection{Swedish}
Swedish\cite{SAG:intro} is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite{H&H:1027}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

Could varandra be thought of as reflexive for 'oss'?
%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Han har inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

unless their in passive form
\enumsentence{
Det äts en fågel på taket \\
A bird is being eaten on the roof}
A very common example of this is sentences with the verbs \emph{finnas},
\emph{saknas} and \emph{fattas}.
\enumsentence{\shortex{3}
{Det & finns & kaffe}
{It & exist & coffee}
{''There is coffee"}}
\enumsentence{\shortex{3}
{Det & saknas&  kaffe}
{It & misses & coffee}
{''There is no coffee"}}
\enumsentence{\shortex{3}
{Det & fattas & kaffe}
{It & misses & coffee}
{''There is no coffee"}}

A special GF category, \verb|SimpleVP|, was needed for verb phrases of the type
described above.

There are also restrictions on the subject.
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågeln på taket. & *Det sitter fågeln på taket.\\
``There are that bird sitting on the roof'' & ``There are the bird sitting on the roof'' \\
\end{tabular}}
The noun must be in indefinite form, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              ``There are some birds on the roof''}
\enumsentence{*Det sitter fåglarna på taket\\
``There are the birds on the roof''}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> CN -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determinened form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definitness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers').

%finnas: här får man ha definit: det finns de som inte vill. Samma RelCl som
%jag just la till, de äppplen. (de + relsats)

%'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'. Kan säga 'en sån katt som sover, med 'sådan' + Rel
%har lagt till Det äpple + Rel (RelCNNP). *Det äpple är rött. men Det äpple jag äter är rött.
%

\subsubsection{Quantifiers, determiners and predeterminers}
The GF analyse differs sometimes. Experiments of how to extract
this information from Talbanken by using Saldo as a starting poin.
One of the biggest differences between the GF analysis and the one made in Talbanken
is related to pronouns. While GF only recognizes the personal pronouns, 
`sådan', `fler`, `somliga' etc. are all regarded as pronouns in Talbanken and Saldo.
In the GF analyse they may be quantifiers, determiners or predeterminers. 
All pronouns from Saldo was listed and by searching Talbanken for which tags
they got we could sort out the ones acting either as
determiners, as `sådana' in sentence \ref{ex:QuantDet}, or as adjectives as
`sådan' in sentence \ref{ex:QuantAdj}. 
\enumsentence{Sådana katter vill jag ha}\label{ex:QuantDet}
\vspace{-3mm}
\enumsentence{Han är sådan}\label{ex:QuantAdj}
\vspace{-3mm}
\enumsentence{Sådant är inte bra}\label{ex:QuantDet2}
Since all determiners can be used alone as noun phrases, sentence
\ref{ex:QuantDet2} is also accepted.
%what is the difference really???
%The techniques of searching Talbanken can be used
%for this purpose as well. 
%combining the information in Saldo and Talbanken
%pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
%has to decide, one category for each?

\subsubsection{Other small things}

\textbf{Genitive} -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form (fixed)
           fixing this include restructuring all???
\textbf{Det-utr}. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

\textbf{finitness}
 'de flesta de gula hästarna'

\textbf{advs}
-- or is this really just AdVs? 
to change position of adv:
normal: after verb 'han äpplet redan'
now want: 'äter redan äpplet'
          'har redan ätit äpplet' 
          'har ätit redan äpplet' :( which doesn't work either :)

\textbf{bara}
but have other words like 'bara' which can be put in other positions.
  but needs to restruct whole of VP, since no field is befor fin.
  Can also act before the noun or as AdV. Examples.
\verb|s SPres Simul Neg Main : han bara sover inte alltid|
  have added a field in VP, a0, show table.
  makes some weird constructions when using modal verbs, 
  \verb|s SPres Anter Pos Main : han bara har alltid sovit|
  but rather than changing to 'han har bara sovit', we keep it. Is a possible
  tolkning and the other case is covered by using it as an AdV, we don't want
  ambiguities.

\textbf{antalet}
Also needed categories for 'pronouns' wich can act in different ways.
and 'antal','sorter' - need indetermined, antal plural, sorter either.
Should be able to modify by a godtyckligt antal adjective, and they must be determined.
'(ett stort antal) människor' and not 'ett stort (antal människor)'
Don't want them as a NPs or CNs since we don't want other Ns to be used this way
'en flicka kaffe', this is then Appos? (Or do we?). Should also not be utsatta for what other
NPs or CN can be. Dessutom, we want to make use of the comp-fiel. 'mamma till', but is this really
the same??

\textbf{VP}
The VP require many changes PPart, bara.
About PPart, but do we need it with new Saldo? 'de är stuckna' , 'den nyligen funna'
for the last one, we need VPSlash to allow adverb.

\textbf{SuperlA}

\subsubsection*{Reflexive pronouns}
%varandra and sin/sitt/sig.
Both sig and sin
Not subject, but NP.
Want the following:
\enumsentence{Han såg sina barns skor.}
\enumsentence{Sina vantar hade han glömt på tåget.}
but not
\enumsentence{*Sina vantar var kvar på tåget.}
\enumsentence{Han ger sina pengar till sina syskon.}
\enumsentence{*Jag ger sina pengar till sina syskon.}
\enumsentence{Han är längre än sin kompis.}
\enumsentence{Hon tyckte om alla sina elever.}
\enumsentence{Han såg sin hund och sitt hus}
\label{ex:refGen}
\enumsentence{Han är längre än sin kompis}
\label{ex:ReflAP}
% hon ber sin katt gå - sin subject in clause.

@@

Kanske mindre om detta, skriv om vad jag nu la till.
om 'de äpplen', which are also used in 'finnas'
Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\
